Uranyl solvation by the three dimensional reference interaction site model

### 2015-02-06 doc/uo2_paper/uo2-II.tex
AM: include figures from cwd
AM: re-generate with elsevier style

%% LyX 2.0.6 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english,twocolumn,showkeys,superscriptaddress]{revtex4}
\usepackage[T1]{fontenc}
\setcounter{secnumdepth}{3}
\usepackage{color}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{esint}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\@ifundefined{textcolor}{}
{%
 \definecolor{BLACK}{gray}{0}
 \definecolor{WHITE}{gray}{1}
 \definecolor{RED}{rgb}{1,0,0}
 \definecolor{GREEN}{rgb}{0,1,0}
 \definecolor{BLUE}{rgb}{0,0,1}
 \definecolor{CYAN}{cmyk}{1,0,0,0}
 \definecolor{MAGENTA}{cmyk}{0,1,0,0}
 \definecolor{YELLOW}{cmyk}{0,0,1,0}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{mathptmx}
\usepackage{times}
\usepackage{microtype}
\def\thetable{\arabic{table}}

@@

\author{Alexei Matveev}

\email{alexei.matveev@gmail.com}

\selectlanguage{english}%

\author{Bo Li}

\affiliation{Department Chemie, Technische Universität München, 85747 Garching,
Germany}

\author{Notker Rösch}

\affiliation{Department Chemie and Catalysis Research Center, Technische Universität
München, 85747 Garching, Germany}

\date{Jan 2015}
\begin{abstract}
We report an implementation of the three-dimensional reference interaction
site model (3D RISM) specifically describing the treatment of the
long-range Coulomb field of charged species represented by point charges
and/or distributed charge density. A comparison of 1D and 3D results
for atomic ions demonstrates the reasonable accuracy even for moderate
box size and grid resolution. In an application to uranyl aqua complexes
with 4\textendash{}6 explicit aqua ligands and an implicit bulk solvent
modeled by RISM, we demonstrate that the 3D technique is not susceptible
to the deficiencies of 1D technique exposed in our previous work {[}Li,
Matveev, Krüger, Rösch, \emph{Comp. Theor. Chem.} \textbf{1051}, p.~151,
(2015){]}. The 3D method eliminates the artificial superposition of
explicit aqua ligands and the RISM medium. It predicts essentially
the same values for uranyl and uranyl-water bond lengths as a state
of the art polarizable continuum model. We find that with the first
solvation shell treated explicitly, the observables are nearly independent
of the closure order. The water exchange activation barrier obtained
in a hybrid approach combining the quantum mechanical description
of the uranyl interaction within the first solvation shell and the
RISM model for the bulk solvent is in close agreement with the experimental
and the best theoretical estimates. 
\end{abstract}

\keywords{Reference interaction site model, aqueous solution, uranyl, solvation}

\maketitle

@@

Thermodynamic and structural properties of the uranyl(VI) cation,
UO$_{2}^{2+}$, the most common uranium species in aqueous solution
are crucial for understanding its environmental chemistry as well
as chemical processes in radioactive waste disposal and recycling
of used nuclear fuel \cite{book:Morss06,book:Grenthe92,Silva95}.
Quantum mechanics (QM) and classical molecular mechanics (MM) simulations
provide the two complimentary ways to address microscopic structure
and energetics of solvated uranyl species. Till now, the most popular
strategy to combine the accuracy of QM methods and the complexity
of the statistical solvent description is to apply a polarizable continuum
model (PCM) to represent the effective electrostatic field of the
solvent, usually employed in combination with phenomenological corrections
for dispersion-repulsion interactions and cavity formation energy
using empiric parameters \cite{Tomasi05,Fuchs02}. The state of the
art PCM is owing its success in part to the conductor-like screening
model (COSMO) of electrostatic interactions \cite{Klamt95} and has
successfully been applied in various studies of uranyl complexes \cite{Fuchs02,Moskaleva04,Cao05,Shamov05,Gutowski06}.
On the other hand, there is a significant effort to improve the empirical
model potentials for MM simulations, such as molecular dynamics (MD)
or Monte-Carlo (MC) methods, that yield the most explicit statistical
solvent description \cite{Guilbaud93,Guilbaud96,Rai12,Pomogaev13,Kerisit13,Tiwari14}.
Though several studies \cite{Gordon00,hagberg05,Frick09,Tirler13}
attempt to combine QM and stochastic MM models such combinations will
inevitably lack the precise reproducibility  and efficiency of analytic
continuum models due to statistical sampling techniques. Yet more
demanding QM dynamic simulations \cite{Buehl05,Buehl06,Buehl11,Nichols08}
are restricted to comparably small numbers of solvent molecules.

Meanwhile, the integral equation theories of molecular liquids \cite{book:Hirata03,book:Hansen06},
may provide sufficient accuracy at an affordable cost for simulating
chemical processes in solution by employing standard numerical methods
on contemporary hardware. Two approaches of this type are the reference
interaction site model (RISM) \cite{Chandler72,Hirata81,Hirata82,Kinoshita97}
and its three-dimensional generalization (3D RISM) \cite{Beglov95,Kovalenko98}
which provide the solvent structure in form of site-site radial distribution
functions (RDFs) or three-dimensional distribution functions of solvent
sites in an external (solute) field, respectively. Both 1D and 3D
theories provide variational expressions for the solvation energy
in the form of the excess chemical potential for several closure types
\cite{Singer85,Kast08} and, hence, intuitive expressions for the
first derivatives. The dielectrically consistent RISM (DRISM) \cite{Perkyns92}
was designed to compensate the error of the gas-like dielectric response
of RISM for electrolytes and ionic solutes \cite{Perkyns94,Chiodo07,Chuev08,Joung13}.

A correct handling of the electrostatic field of a charge distribution
requires additional care when periodic boundary conditions are employed
to simulate finite systems \cite{Makov95}. It was understood early
that the long-range asymptotes of the radial correlation functions
induced by Coulomb interactions in RISM need a separate treatment
as well \cite{Ng74}. For 3D RISM the few practical approaches include
the background charge correction technique \cite{Kovalenko00b}, monopole
correction \cite{Kloss08b}, and an exact re-summation for a superposition
of gaussian charges \cite{Perkyns10}. In this work we present a generalization
of the exact re-summation technique for arbitrary solute charge densities.

The integral equation theory has been combined with Hartree--Fock
electronic structure theory for self-consistent field calculations
(RISM SCF) in the pioneering work of Ten-no, Hirata, and Kato and
has been first applied to formaldehyde and carbonyl compounds in water
\cite{Tenno93,Tenno94}. An extension to 3D RISM SCF was developed
for density functional theory (DFT) \cite{Kovalenko99} and \emph{ab
initio} calculations \cite{Kloss08}. The analytical gradients for
the self-consistent combination of 3D RISM and DFT \cite{Gusarov06}
simplify explorations of free energy landscapes. Some of the recent
developments of hybrid approaches combining integral equation theories
and quantum chemistry are discussed in the recent review by Sato \cite{Sato13}.

@@

In the 3D RISM approach, the solvent sites are exposed to an external
field of a solute of general shape. The solute-solvent interactions
are thus assumed to be pairwise, which in this formulation excludes
\emph{polarizable} solutes. Further, the interactions are usually
approximated by superposition of pairwise site-site contributions.
The 3D RISM equations operating with site-specific distributions can
be written in the following form \cite{book:Hirata03}:

@@

The solvent susceptibility $\chi$ in the first of the three Eqs.~\ref{eq:rsim-h-c-t}
is represented by its (square symmetric) matrix $\chi_{\alpha\gamma}(r)$
derived from the radial site-site correlation functions of the pure
solvent in 1D RISM equations for a specific temperature and solvent
number density $\rho$. \cite{book:Hirata03}. The star denotes the
convolution integral and the sum over corresponding matrix indices.
For that reason it is convenient to represent the solvent susceptibility
by its (dimensionless) Fourier image $\tilde{\chi}_{\alpha\gamma}(k)$
\cite{Chandler72,book:Hirata03}. The solvent susceptibility as a
function of either $r$ or $k$ is spherically symmetric by construction.

The second of the Eqs.~\ref{eq:rsim-h-c-t} introduces the indirect
correlation $t$ that is conveniently used in the closure relation
--- the third and last of Eqs.~\ref{eq:rsim-h-c-t}. Closure is a
local relation between the values of site potential $v_{\alpha}(\vec{r})$
and, in this case, the indirect correlation $t_{\alpha}(\vec{r})$
at each point $\vec{r}$ and the value of the direct correlation $h_{\alpha}(\vec{r})$
at this point. It is the only non-linear relation between the unknowns
$h$, $c$ and $t$ in Eqs.~\ref{eq:rsim-h-c-t}. The site-specific
potentials $v$ are measured in units of temperature and thus are
also dimensionless. For the purposes of this work we have chosen a
very specific, though popular, form of the closure which happens to
be a relation between $h$ and the sum $x=-v+t$ sometimes referred
to as a renormalized indirect correlation \cite{Kast08}:
\begin{equation}
f(x)=\begin{cases}
\mathrm{e}^{x}-1 & x\le0\\
\sum_{k=1}^{n}x^{k}/k! & x>0
\end{cases}\label{eq:closure-pse}
\end{equation}
In the limit of $n\to\infty$ one recovers the hypernetted chain (HNC)
closure. Another popular closure is obtained for $n=1$. It was proposed
by Kovalenko and Hirata (KH) for improving the numerical stability
of the equations \cite{Kovalenko99}. Higher order closures with $n>1$
are referred to as partial series expansion closures of order $n$
(PSE$n$). For consistency, we will also refer to the KH closure as
PSE1 in this work.

We used pair-wise interactions between solute and solvent sites $i$
and $\alpha$ to define the solute field: $v_{\alpha}(\vec{r})=\beta\sum_{i}u_{i\alpha}(|\vec{r}-\vec{r}_{i}|)$
where $\beta$ is the inverse temperature. Each potential $u_{i\alpha}$
is defined as the sum of the Coulomb interaction $u_{i\alpha}^{\text{C}}(r)=q_{i}q_{\alpha}a[f_{\text{CS}}(ar)+f_{\text{CL}}(ar)]$,
where we separated the singular short-range $f_{\text{CS}}(r)=\mathrm{erfc}(r)/r$
from the regular long-range $f_{\text{CL}}(r)=\mathrm{erf}(r)/r$
part, and a Lennard--Jones (LJ) type potential, $u_{i\alpha}^{\text{LJ}}(r)=\epsilon_{i\alpha}f_{\text{LJ}}(r/\sigma_{i\alpha})$
with a LJ form $f_{\text{LJ}}(r)=4(r^{-12}-r^{-6})$. Parameters of
the pair interactions were derived from site parameters using the
Lorentz--Berthelot rules \cite{book:Hirata03}. The Ewald range parameter
$a=1.2$~Å$^{-1}$ is common for all site pairs. Further, to circumvent
numerical problems with singularities the primitive forms $f_{\text{CS}}(r)$
and $f_{\text{LJ}}(r)$ were regularized by substituting them with
quadratic functions $\bar{f}(r)=f(b)+f^{\prime}(b)(r^{2}-b^{2})/2b$
for $r<b$. For our systems at room temperature the results were almost
indistinguishable for finite $b\le0.5$ so that we settled with $b=0.2$
in this work.

@@

The analytical expression for the excess chemical potential $\mu$
of an unstructured solute and the HNC closure has been derived by
Singer and Chandler (here in temperature units) \cite{Singer85}:
\begin{equation}
\mu_{\text{HNC}}=\rho\int d^{3}r\sum_{\alpha}\left[h_{\alpha}^{2}/2-c_{\alpha}-c_{\alpha}h_{\alpha}/2\right].\label{eq:chem-pot-hnc}
\end{equation}
A modification for the KH closure \cite{Kovalenko99} and a generalization
for the arbitrary closure of the form $h=f(x)$ with $x=-v+t$ are
available: 
\begin{equation}
\mu=\mu_{\text{HNC}}+\rho\int\sum_{\alpha}\phi(x_{\alpha})d^{3}r
\end{equation}
with $\phi(x)=f(x)-x-\int_{0}^{x}f(y)dy$ having particularly a simple
form for the PSE$n$ closures: $\phi_{n}(x)=-x^{n+1}/(n+1)!$ for
$x>0$ and zero otherwise \cite{Kast08}. In the case of the KH (PSE1)
closure the additional term $-x^{2}/2$ cancels the $h^{2}/2$ term
of the HNC integrand, Eq.~\ref{eq:chem-pot-hnc}, in the regions
where $x=h>0$.

The analytical expression for the excess chemical potential is a path-independent
integral of the work required to ``switch on'' the external (solute)
field $v$ \cite{Singer85,Kast08}. This property leads to straightforward
expressions for derivatives of the chemical potential with respect
to parameters affecting the external field, such as the location of
solute sites. Such a derivative is an expectation value of the corresponding
force, $d\mu/ds=\int(dv/ds)gd^{3}r$, with $g=h+1$.

The free energy surface as a function of solute parameters $s$ is
constructed as a composite functional
\begin{equation}
G[n_{\text{e}},s]=E[n_{\text{e}},s]+\mu(s),
\end{equation}
where $E[n_{\text{e}},s]$ is the solute internal energy depending
on the solutes' geometric degrees of freedom $s$ and, in the case
of a QM solute, also on the electronic density $n_{\text{e}}$, but
not on the solvent site distribution functions $h$ \cite{Bo15}.

The uranyl solvation energy is then estimated as\textcolor{green}{{}
}
\begin{eqnarray}
\Delta G_{\text{sol}} & = & G[\text{UO}_{2}(\text{H}_{2}\text{O})_{n}^{2+}]-nG[\text{H}_{2}\text{O}]-E[\text{UO}_{2}^{2+}]\label{eq:dG-sol-definition}
\end{eqnarray}
for $n\ge0$ explicit aqua ligands. Note that without proper statistical
sampling over the solute degrees of freedom $s$ this approximation
of the solvation energy, justified to some extent for sufficiently
rigid solutes, breaks for increasingly many explicit water molecules
outside of the tightly bound first solvation shell as $n\to\infty$.
Indeed, in a gedankenexperiment the difference $G[(\text{H}_{2}\text{O})_{n}]-nG[\text{H}_{2}\text{O}]$
for the above definition of $G$ applied to water droplets asymptotically
approaches $n(\bar{E}[\text{H}_{2}\text{O}]-G[\text{H}_{2}\text{O}])$
where $\bar{E}$ is the average configurational energy of a water
molecule which is about $-$9.9~kcal/mol \cite{Berendsen87,Mahoney00}
to be compared to the solvation energy of $-$6.3~kcal/mol derived
from the ratio of the equilibrium vapor/liquid concentrations \cite{Bridgeman64}.

@@

Because of the highly non-linear closure relation, a single variable
fix point iteration scheme, $t^{(n+1)}=T[t^{(n)}]$, with $T[t]=(\chi-1)\ast c(t)$
and $c(t)=f(-v+t)-t$ as derived from the non-linear equation system,
Eqs.~\ref{eq:rsim-h-c-t}, is bound to diverge for all but the simplest
cases. To overcome the divergence problem we solve the reduced non-linear
problem $F[t]=T[t]-t=0$ using the Newton--Krylov non-linear solver
\cite{petsc-user-ref} with the analytic matrix-free Jacobian, $J[t]=\delta F/\delta t$,
occasionally preceded by a few Picard iterations.

The long-range field of a charged solute requires a special treatment
in the 3D case. All of the practical approaches, including the Ng
scheme \cite{Ng74}, separate the smooth asymptote $v_{\text{L}}$
of the site specific potential off the indirect correlation $t$ as
well as $-v_{\text{L}}$ off the direct correlation $c$ to operate
primarily with the short range counterparts $v_{\text{S}}$ , $c_{\text{S}}$,
and $t_{\text{S}}$. An iteration scheme for the new primary variable
$t_{\text{S}}$ (and the resulting non-linear problem) will look slightly
different, though:

@@

Indeed, the low $k$ behavior of another convolution operand, the
solvent susceptibility is regular 
\begin{equation}
\tilde{\chi}_{\alpha\gamma}(k)\sim\rho\kappa+\frac{1}{2}d_{\alpha\gamma}k^{2}
\end{equation}
with the limit for $k\to0$ being proportional to the solvent compressibility
$\kappa$ (here in temperature units) \cite{book:Hansen06}. More
importantly, this limit is the same for all site pairs $\alpha$ and
$\gamma$. Though not enforced, this property is obeyed in practical
calculations with reasonable accuracy. The second, $o(k^{2})$, term
is site pair specific and is in particular responsible for the dielectric
response of the solvent --- it is this term that is addressed by the
dielectric correction of the DRISM approach \cite{Perkyns92,Cummings81,Kovalenko00b}.
The result of the convolution, 
\begin{equation}
\tilde{\tau}_{\alpha}(\vec{k})=\tilde{n}(\vec{k})\left[\tilde{\varphi}(k)\sum_{\gamma}\tilde{\chi}_{\alpha\gamma}(k)q_{\gamma}\right]\sim2\pi Q\sum_{\gamma}d_{\alpha\gamma}q_{\gamma}\label{eq:tau}
\end{equation}
is regular assuming net neutrality of the solvent, $\sum_{\gamma}q_{\gamma}=0$,
which makes the compressibility term vanish. Note that the terms in
square brackets, $\tilde{\varphi}(k)\sum_{\gamma}\tilde{\chi}_{\alpha\gamma}(k)q_{\gamma}=:\tilde{\theta}_{\alpha}(k)$
(one per solvent site $\alpha$), are convolutions of spherically
symmetric functions neither of which is solute specific. The functions
$\tilde{\theta}_{\alpha}(k)$ are site-specific solvent properties
representing response to a unit (here gaussian-broadened) charge that
should be computed using 1D techniques, just as the solvent susceptibility
$\tilde{\chi}_{\alpha\gamma}(k)$ itself. In the 1D case our deliberate
choice of the self-inverse discrete sine transform of type IV (DST-IV)
as a basic operation eliminates the need to treat the singularities
at $r=0$ and $k=0$ as the discrete representation do not include
these points by construction \cite{FFTW05}. The solute part is the
electric form factor $\tilde{n}(\vec{k})$ of an arbitrary solute
charge distribution $n(\vec{r})$ with a low $k$ expansion including
all multipoles of the charge distribution; it is treated by regular
3D techniques.

>>>>

%% LyX 2.0.6 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english,5p,sort&compress,times]{elsarticle}
\usepackage[T1]{fontenc}
\usepackage{color}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{esint}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
%\usepackage{mathptmx}
%\usepackage{times}
%\def\thetable{\arabic{table}}
\usepackage{microtype}
\journal{Computational and Theoretical Chemistry}

@@

\author[tum]{Alexei Matveev\corref{corr}}

\ead{alexei.matveev@gmail.com}

\author[tum]{Bo Li}

\ead{fixme@abc.com}

\author[tum,crc]{Notker Rösch}

\ead{fixme@xyz.com}

\cortext[corr]{Corresponding author}

\address[tum]{Department Chemie, Technische Universität München, 85747 Garching,
Germany}

\address[crc]{Catalysis Research Center, Technische Universität München, 85747
Garching, Germany}
\begin{abstract}
We report an implementation of the three-dimensional reference interaction
site model (3D RISM) specifically describing the treatment of the
long-range Coulomb field of charged species represented by point charges
and/or distributed charge density. A comparison of 1D and 3D results
for atomic ions demonstrates the reasonable accuracy even for moderate
box size and grid resolution. In an application to uranyl aqua complexes
with 4\textendash{}6 explicit aqua ligands and an implicit bulk solvent
modeled by RISM, we demonstrate that the 3D technique is not susceptible
to the deficiencies of 1D technique exposed in our previous work {[}Li,
Matveev, Krüger, Rösch, \emph{Comp. Theor. Chem.} \textbf{1051}, p.~151,
(2015){]}. The 3D method eliminates the artificial superposition of
explicit aqua ligands and the RISM medium. It predicts essentially
the same values for uranyl and uranyl-water bond lengths as a state
of the art polarizable continuum model. We find that with the first
solvation shell treated explicitly, the observables are nearly independent
of the closure order. The water exchange activation barrier obtained
in a hybrid approach combining the quantum mechanical description
of the uranyl interaction within the first solvation shell and the
RISM model for the bulk solvent is in close agreement with the experimental
and the best theoretical estimates. \end{abstract}
\begin{keyword}
Reference interaction site model, uranyl, aqueous solution
\end{keyword}
\maketitle

@@

Thermodynamic and structural properties of the uranyl(VI) cation,
UO$_{2}^{2+}$, the most common uranium species in aqueous solution
are crucial for understanding its environmental chemistry as well
as chemical processes in radioactive waste disposal and recycling
of used nuclear fuel \citep{book:Morss06,book:Grenthe92,Silva95}.
Quantum mechanics (QM) and classical molecular mechanics (MM) simulations
provide the two complimentary ways to address microscopic structure
and energetics of solvated uranyl species. Till now, the most popular
strategy to combine the accuracy of QM methods and the complexity
of the statistical solvent description is to apply a polarizable continuum
model (PCM) to represent the effective electrostatic field of the
solvent, usually employed in combination with phenomenological corrections
for dispersion-repulsion interactions and cavity formation energy
using empiric parameters \citep{Tomasi05,Fuchs02}. The state of the
art PCM is owing its success in part to the conductor-like screening
model (COSMO) of electrostatic interactions \citep{Klamt95} and has
successfully been applied in various studies of uranyl complexes \citep{Fuchs02,Moskaleva04,Cao05,Shamov05,Gutowski06}.
On the other hand, there is a significant effort to improve the empirical
model potentials for MM simulations, such as molecular dynamics (MD)
or Monte-Carlo (MC) methods, that yield the most explicit statistical
solvent description \citep{Guilbaud93,Guilbaud96,Rai12,Pomogaev13,Kerisit13,Tiwari14}.
Though several studies \citep{Gordon00,hagberg05,Frick09,Tirler13}
attempt to combine QM and stochastic MM models such combinations will
inevitably lack the precise reproducibility  and efficiency of analytic
continuum models due to statistical sampling techniques. Yet more
demanding QM dynamic simulations \citep{Buehl05,Buehl06,Buehl11,Nichols08}
are restricted to comparably small numbers of solvent molecules.

Meanwhile, the integral equation theories of molecular liquids \citep{book:Hirata03,book:Hansen06},
may provide sufficient accuracy at an affordable cost for simulating
chemical processes in solution by employing standard numerical methods
on contemporary hardware. Two approaches of this type are the reference
interaction site model (RISM) \citep{Chandler72,Hirata81,Hirata82,Kinoshita97}
and its three-dimensional generalization (3D RISM) \citep{Beglov95,Kovalenko98}
which provide the solvent structure in form of site-site radial distribution
functions (RDFs) or three-dimensional distribution functions of solvent
sites in an external (solute) field, respectively. Both 1D and 3D
theories provide variational expressions for the solvation energy
in the form of the excess chemical potential for several closure types
\citep{Singer85,Kast08} and, hence, intuitive expressions for the
first derivatives. The dielectrically consistent RISM (DRISM) \citep{Perkyns92}
was designed to compensate the error of the gas-like dielectric response
of RISM for electrolytes and ionic solutes \citep{Perkyns94,Chiodo07,Chuev08,Joung13}.

A correct handling of the electrostatic field of a charge distribution
requires additional care when periodic boundary conditions are employed
to simulate finite systems \citep{Makov95}. It was understood early
that the long-range asymptotes of the radial correlation functions
induced by Coulomb interactions in RISM need a separate treatment
as well \citep{Ng74}. For 3D RISM the few practical approaches include
the background charge correction technique \citep{Kovalenko00b},
monopole correction \citep{Kloss08b}, and an exact re-summation for
a superposition of gaussian charges \citep{Perkyns10}. In this work
we present a generalization of the exact re-summation technique for
arbitrary solute charge densities.

The integral equation theory has been combined with Hartree--Fock
electronic structure theory for self-consistent field calculations
(RISM SCF) in the pioneering work of Ten-no, Hirata, and Kato and
has been first applied to formaldehyde and carbonyl compounds in water
\citep{Tenno93,Tenno94}. An extension to 3D RISM SCF was developed
for density functional theory (DFT) \citep{Kovalenko99} and \emph{ab
initio} calculations \citep{Kloss08}. The analytical gradients for
the self-consistent combination of 3D RISM and DFT \citep{Gusarov06}
simplify explorations of free energy landscapes. Some of the recent
developments of hybrid approaches combining integral equation theories
and quantum chemistry are discussed in the recent review by Sato \citep{Sato13}.

@@

In the 3D RISM approach, the solvent sites are exposed to an external
field of a solute of general shape. The solute-solvent interactions
are thus assumed to be pairwise, which in this formulation excludes
\emph{polarizable} solutes. Further, the interactions are usually
approximated by superposition of pairwise site-site contributions.
The 3D RISM equations operating with site-specific distributions can
be written in the following form \citep{book:Hirata03}:

@@

The solvent susceptibility $\chi$ in the first of the three Eqs.~\ref{eq:rsim-h-c-t}
is represented by its (square symmetric) matrix $\chi_{\alpha\gamma}(r)$
derived from the radial site-site correlation functions of the pure
solvent in 1D RISM equations for a specific temperature and solvent
number density $\rho$. \citep{book:Hirata03}. The star denotes the
convolution integral and the sum over corresponding matrix indices.
For that reason it is convenient to represent the solvent susceptibility
by its (dimensionless) Fourier image $\tilde{\chi}_{\alpha\gamma}(k)$
\citep{Chandler72,book:Hirata03}. The solvent susceptibility as a
function of either $r$ or $k$ is spherically symmetric by construction.

The second of the Eqs.~\ref{eq:rsim-h-c-t} introduces the indirect
correlation $t$ that is conveniently used in the closure relation
--- the third and last of Eqs.~\ref{eq:rsim-h-c-t}. Closure is a
local relation between the values of site potential $v_{\alpha}(\vec{r})$
and, in this case, the indirect correlation $t_{\alpha}(\vec{r})$
at each point $\vec{r}$ and the value of the direct correlation $h_{\alpha}(\vec{r})$
at this point. It is the only non-linear relation between the unknowns
$h$, $c$ and $t$ in Eqs.~\ref{eq:rsim-h-c-t}. The site-specific
potentials $v$ are measured in units of temperature and thus are
also dimensionless. For the purposes of this work we have chosen a
very specific, though popular, form of the closure which happens to
be a relation between $h$ and the sum $x=-v+t$ sometimes referred
to as a renormalized indirect correlation \citep{Kast08}:
\begin{equation}
f(x)=\begin{cases}
\mathrm{e}^{x}-1 & x\le0\\
\sum_{k=1}^{n}x^{k}/k! & x>0
\end{cases}\label{eq:closure-pse}
\end{equation}
In the limit of $n\to\infty$ one recovers the hypernetted chain (HNC)
closure. Another popular closure is obtained for $n=1$. It was proposed
by Kovalenko and Hirata (KH) for improving the numerical stability
of the equations \citep{Kovalenko99}. Higher order closures with
$n>1$ are referred to as partial series expansion closures of order
$n$ (PSE$n$). For consistency, we will also refer to the KH closure
as PSE1 in this work.

We used pair-wise interactions between solute and solvent sites $i$
and $\alpha$ to define the solute field: $v_{\alpha}(\vec{r})=\beta\sum_{i}u_{i\alpha}(|\vec{r}-\vec{r}_{i}|)$
where $\beta$ is the inverse temperature. Each potential $u_{i\alpha}$
is defined as the sum of the Coulomb interaction $u_{i\alpha}^{\text{C}}(r)=q_{i}q_{\alpha}a[f_{\text{CS}}(ar)+f_{\text{CL}}(ar)]$,
where we separated the singular short-range $f_{\text{CS}}(r)=\mathrm{erfc}(r)/r$
from the regular long-range $f_{\text{CL}}(r)=\mathrm{erf}(r)/r$
part, and a Lennard--Jones (LJ) type potential, $u_{i\alpha}^{\text{LJ}}(r)=\epsilon_{i\alpha}f_{\text{LJ}}(r/\sigma_{i\alpha})$
with a LJ form $f_{\text{LJ}}(r)=4(r^{-12}-r^{-6})$. Parameters of
the pair interactions were derived from site parameters using the
Lorentz--Berthelot rules \citep{book:Hirata03}. The Ewald range parameter
$a=1.2$~Å$^{-1}$ is common for all site pairs. Further, to circumvent
numerical problems with singularities the primitive forms $f_{\text{CS}}(r)$
and $f_{\text{LJ}}(r)$ were regularized by substituting them with
quadratic functions $\bar{f}(r)=f(b)+f^{\prime}(b)(r^{2}-b^{2})/2b$
for $r<b$. For our systems at room temperature the results were almost
indistinguishable for finite $b\le0.5$ so that we settled with $b=0.2$
in this work.

@@

The analytical expression for the excess chemical potential $\mu$
of an unstructured solute and the HNC closure has been derived by
Singer and Chandler (here in temperature units) \citep{Singer85}:
\begin{equation}
\mu_{\text{HNC}}=\rho\int d^{3}r\sum_{\alpha}\left[h_{\alpha}^{2}/2-c_{\alpha}-c_{\alpha}h_{\alpha}/2\right].\label{eq:chem-pot-hnc}
\end{equation}
A modification for the KH closure \citep{Kovalenko99} and a generalization
for the arbitrary closure of the form $h=f(x)$ with $x=-v+t$ are
available: 
\begin{equation}
\mu=\mu_{\text{HNC}}+\rho\int\sum_{\alpha}\phi(x_{\alpha})d^{3}r
\end{equation}
with $\phi(x)=f(x)-x-\int_{0}^{x}f(y)dy$ having particularly a simple
form for the PSE$n$ closures: $\phi_{n}(x)=-x^{n+1}/(n+1)!$ for
$x>0$ and zero otherwise \citep{Kast08}. In the case of the KH (PSE1)
closure the additional term $-x^{2}/2$ cancels the $h^{2}/2$ term
of the HNC integrand, Eq.~\ref{eq:chem-pot-hnc}, in the regions
where $x=h>0$.

The analytical expression for the excess chemical potential is a path-independent
integral of the work required to ``switch on'' the external (solute)
field $v$ \citep{Singer85,Kast08}. This property leads to straightforward
expressions for derivatives of the chemical potential with respect
to parameters affecting the external field, such as the location of
solute sites. Such a derivative is an expectation value of the corresponding
force, $d\mu/ds=\int(dv/ds)gd^{3}r$, with $g=h+1$.

The free energy surface as a function of solute parameters $s$ is
constructed as a composite functional
\begin{equation}
G[n_{\text{e}},s]=E[n_{\text{e}},s]+\mu(s),
\end{equation}
where $E[n_{\text{e}},s]$ is the solute internal energy depending
on the solutes' geometric degrees of freedom $s$ and, in the case
of a QM solute, also on the electronic density $n_{\text{e}}$, but
not on the solvent site distribution functions $h$ \citep{Bo15}.

The uranyl solvation energy is then estimated as\textcolor{green}{{}
}
\begin{eqnarray}
\Delta G_{\text{sol}} & = & G[\text{UO}_{2}(\text{H}_{2}\text{O})_{n}^{2+}]-nG[\text{H}_{2}\text{O}]-E[\text{UO}_{2}^{2+}]\label{eq:dG-sol-definition}
\end{eqnarray}
for $n\ge0$ explicit aqua ligands. Note that without proper statistical
sampling over the solute degrees of freedom $s$ this approximation
of the solvation energy, justified to some extent for sufficiently
rigid solutes, breaks for increasingly many explicit water molecules
outside of the tightly bound first solvation shell as $n\to\infty$.
Indeed, in a gedankenexperiment the difference $G[(\text{H}_{2}\text{O})_{n}]-nG[\text{H}_{2}\text{O}]$
for the above definition of $G$ applied to water droplets asymptotically
approaches $n(\bar{E}[\text{H}_{2}\text{O}]-G[\text{H}_{2}\text{O}])$
where $\bar{E}$ is the average configurational energy of a water
molecule which is about $-$9.9~kcal/mol \citep{Berendsen87,Mahoney00}
to be compared to the solvation energy of $-$6.3~kcal/mol derived
from the ratio of the equilibrium vapor/liquid concentrations \citep{Bridgeman64}.

@@

Because of the highly non-linear closure relation, a single variable
fix point iteration scheme, $t^{(n+1)}=T[t^{(n)}]$, with $T[t]=(\chi-1)\ast c(t)$
and $c(t)=f(-v+t)-t$ as derived from the non-linear equation system,
Eqs.~\ref{eq:rsim-h-c-t}, is bound to diverge for all but the simplest
cases. To overcome the divergence problem we solve the reduced non-linear
problem $F[t]=T[t]-t=0$ using the Newton--Krylov non-linear solver
\citep{petsc-user-ref} with the analytic matrix-free Jacobian, $J[t]=\delta F/\delta t$,
occasionally preceded by a few Picard iterations.

The long-range field of a charged solute requires a special treatment
in the 3D case. All of the practical approaches, including the Ng
scheme \citep{Ng74}, separate the smooth asymptote $v_{\text{L}}$
of the site specific potential off the indirect correlation $t$ as
well as $-v_{\text{L}}$ off the direct correlation $c$ to operate
primarily with the short range counterparts $v_{\text{S}}$ , $c_{\text{S}}$,
and $t_{\text{S}}$. An iteration scheme for the new primary variable
$t_{\text{S}}$ (and the resulting non-linear problem) will look slightly
different, though:

@@

Indeed, the low $k$ behavior of another convolution operand, the
solvent susceptibility is regular 
\begin{equation}
\tilde{\chi}_{\alpha\gamma}(k)\sim\rho\kappa+\frac{1}{2}d_{\alpha\gamma}k^{2}
\end{equation}
with the limit for $k\to0$ being proportional to the solvent compressibility
$\kappa$ (here in temperature units) \citep{book:Hansen06}. More
importantly, this limit is the same for all site pairs $\alpha$ and
$\gamma$. Though not enforced, this property is obeyed in practical
calculations with reasonable accuracy. The second, $o(k^{2})$, term
is site pair specific and is in particular responsible for the dielectric
response of the solvent --- it is this term that is addressed by the
dielectric correction of the DRISM approach \citep{Perkyns92,Cummings81,Kovalenko00b}.
The result of the convolution, 
\begin{equation}
\tilde{\tau}_{\alpha}(\vec{k})=\tilde{n}(\vec{k})\left[\tilde{\varphi}(k)\sum_{\gamma}\tilde{\chi}_{\alpha\gamma}(k)q_{\gamma}\right]\sim2\pi Q\sum_{\gamma}d_{\alpha\gamma}q_{\gamma}\label{eq:tau}
\end{equation}
is regular assuming net neutrality of the solvent, $\sum_{\gamma}q_{\gamma}=0$,
which makes the compressibility term vanish. Note that the terms in
square brackets, $\tilde{\varphi}(k)\sum_{\gamma}\tilde{\chi}_{\alpha\gamma}(k)q_{\gamma}=:\tilde{\theta}_{\alpha}(k)$
(one per solvent site $\alpha$), are convolutions of spherically
symmetric functions neither of which is solute specific. The functions
$\tilde{\theta}_{\alpha}(k)$ are site-specific solvent properties
representing response to a unit (here gaussian-broadened) charge that
should be computed using 1D techniques, just as the solvent susceptibility
$\tilde{\chi}_{\alpha\gamma}(k)$ itself. In the 1D case our deliberate
choice of the self-inverse discrete sine transform of type IV (DST-IV)
as a basic operation eliminates the need to treat the singularities
at $r=0$ and $k=0$ as the discrete representation do not include
these points by construction \citep{FFTW05}. The solute part is the
electric form factor $\tilde{n}(\vec{k})$ of an arbitrary solute
charge distribution $n(\vec{r})$ with a low $k$ expansion including
all multipoles of the charge distribution; it is treated by regular
3D techniques.
